---
title: "EV data findings"
author: "pablo paulsen"
date: "23/11/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(fig.width=9, fig.height=5) 
library(dplyr)
library(ggplot2)
library(tidyr)
library(quantmod)
library(forecast)
library(tseries)
```

```{r functions}
HDD = function (data, temp = 16) {
  mon_sum = data.frame(Year = NA, Month = NA, HDD = NA)[-1,]
  for (y in min(data$Year):max(data$Year)){
    for (m in min(data[data$Year == y,]$Month):max(data[data$Year == y,]$Month)) {
      #print(paste(y," ", m))
      adj_temps = -data[data$Year == y & data$Month == m,]$Temp+temp
      #print(sum(adj_temps*(adj_temps > 0))/24)
      mon_sum[nrow(mon_sum) + 1,] = c(y,m, sum(adj_temps*(adj_temps > 0))/24)
    }
  }
  mon_sum$City = data$City[1]
  return(mon_sum)
}

CDD = function (data, temp = 22) {
  mon_sum = data.frame(Year = NA, Month = NA, CDD = NA)[-1,]
  for (y in min(data$Year):max(data$Year)){
    for (m in min(data[data$Year == y,]$Month):max(data[data$Year == y,]$Month)) {
      #print(paste(y," ", m))
      adj_temps = data[data$Year == y & data$Month == m,]$Temp-temp
      #print(sum(adj_temps*(adj_temps > 0))/24)
      mon_sum[nrow(mon_sum) + 1,] = c(y,m, sum(adj_temps*(adj_temps > 0))/24)

    }
  }
  mon_sum$City = data$City[1]

  return(mon_sum)
}
avg_temp = function(data) {
  avg_temp = data %>% 
    group_by(Year, Month, City) %>% 
    summarise(avg_temp = mean(Temp))
  return(avg_temp)
}

yearly_line = function(period = 12, count = 10) {
  for (i in 0:count) {
    abline(v = i*period+1, lty = 3)
  }
}
```


```{r data_setup, message = FALSE}
month_length = c(31,28,31,30,31,30,31,31,30,31,30,31)
#list index (first) is the location weather to use and value (second) is the region in the data 
weather_regions = list("Auckland" = "Auckland", "Upper Hutt" = "Wellington","Christchurch" = "Christchurch", "Dunedin" = "Coastal Otago", "Hamilton" = "Waikato", "Rotorua" = "Bay of Plenty","Christchurch" = "Mid Canterbury", "Christchurch" = "North Canterbury", "Christchurch" = "South Canterbury", "Clyde" = "Central Otago", "Nelson" = "Nelson", "Auckland" = "Whangarei", "Auckland" = "Far North", "Upper Hutt" = "Wairarapa", "Auckland" = "Rodney", "Dunedin" = "Waitaki", "Nelson" = "Golden Bay",  "Auckland" = "Coromandel", "Palmerston North" = "Manawatu", "Nelson" = "Marlborough", "Stratford" = "Taranaki", "Napier" = "Hawkes Bay", "Invercargill" = "Southland")



#main EV data
EV_data = read.csv('../Ftf Efficiency Dataset/ftf_ev_efficiency_distance_models_20211006_v1.0.csv', stringsAsFactors = T)[-1]
EV_data$weather_region = as.factor(names(weather_regions)[match(EV_data$region,  weather_regions)])

#sort factor by popularity
region_pop = EV_data %>%
  group_by(weather_region) %>% 
  summarise(count =  n_distinct(vehicle)) %>% 
  arrange(-count)
model_pop = EV_data %>%
  group_by(model) %>% 
  summarise(count =  n_distinct(vehicle)) %>% 
  arrange(-count)
EV_data$weather_region = factor(EV_data$weather_region, levels = region_pop$weather_region)
EV_data$model = factor(EV_data$model, levels = model_pop$model)

#remove PHEV
EV_data = EV_data[EV_data$model != "Mitsubishi Outlander" & EV_data$model != "Toyota Prius" & EV_data$model != "Mini Countryman PHEV", ]



#weather data
auckland_weather = read.csv('weather_data/auckland_motat_ews_data.csv',na.strings = "-",stringsAsFactors = T)
auckland_weather$Month = match(auckland_weather$Month, month.abb)
#Wellington does not have NIWA stations so closest is upperhutt
upperhutt_weather = read.csv('weather_data/upperhutt_trentham_ews_data.csv',na.strings = "-",stringsAsFactors = T)
upperhutt_weather$Month = match(upperhutt_weather$Month, month.abb)
christchurch_weather = read.csv('weather_data/christchurch_kyle_st_ews_data.csv',na.strings = "-", stringsAsFactors = T)
christchurch_weather$Month = match(christchurch_weather$Month, month.abb)
#use Dunedin weather for coastal Otago
dunedin_weather = read.csv('weather_data/dunedin_musselburgh_ews_data.csv',na.strings = "-",stringsAsFactors = T)
dunedin_weather$Month = match(dunedin_weather$Month, month.abb)
#use hamilton weather for Waikato
hamilton_weather = read.csv('weather_data/hamilton_ruakura_ews_data.csv', na.strings = "-", stringsAsFactors = T)
hamilton_weather$Month = match(hamilton_weather$Month, month.abb)
#use rotorua for bay of plenty
rotorua_weather = read.csv('weather_data/rotorua_ews_data.csv', na.strings = "-", stringsAsFactors = T)
rotorua_weather$Month = match(rotorua_weather$Month, month.abb)
#use clyde for central Otago
clyde_weather = read.csv('weather_data/clyde_ews_data.csv', na.strings = "-", stringsAsFactors = T)
clyde_weather$Month = match(clyde_weather$Month, month.abb)
nelson_weather = read.csv('weather_data/nelson_ews_data.csv', na.strings = "-", stringsAsFactors = T)
nelson_weather$Month = match(nelson_weather$Month, month.abb)
#use palmerston north for Manawatu
palmerston_weather = read.csv('weather_data/palmerston_north_ews_data.csv', na.strings = "-", stringsAsFactors = T)
palmerston_weather$Month = match(palmerston_weather$Month, month.abb)
#use stratford for Taranaki
stratford_weather = read.csv('weather_data/stratford_ews_data.csv', na.strings = "-", stringsAsFactors = T)
stratford_weather$Month = match(stratford_weather$Month, month.abb)
#use napier for Hawkes Bay
napier_weather = read.csv('weather_data/napier_ews_data.csv', na.strings = "-", stringsAsFactors = T)
napier_weather$Month = match(napier_weather$Month, month.abb)
#use invercargill for southland
invercargill_weather = read.csv('weather_data/invercargill_ews_data.csv', na.strings = "-", stringsAsFactors = T)
invercargill_weather$Month = match(invercargill_weather$Month, month.abb)

#temp calculations
HDD_data = Reduce(rbind, list(HDD(auckland_weather), HDD(upperhutt_weather), HDD(christchurch_weather),
                HDD(dunedin_weather), HDD(hamilton_weather), HDD(rotorua_weather), HDD(clyde_weather),
                HDD(nelson_weather), HDD(palmerston_weather), HDD(stratford_weather), HDD(napier_weather),
                HDD(invercargill_weather)))
CDD_data = Reduce(rbind, list(CDD(auckland_weather), CDD(upperhutt_weather), CDD(christchurch_weather),
                CDD(dunedin_weather), CDD(hamilton_weather), CDD(rotorua_weather), CDD(clyde_weather),
                CDD(nelson_weather), CDD(palmerston_weather), CDD(stratford_weather), CDD(napier_weather),
                CDD(invercargill_weather)))
avg_temp_data = Reduce(rbind, list(avg_temp(auckland_weather), avg_temp(upperhutt_weather), avg_temp(christchurch_weather),
                avg_temp(dunedin_weather), avg_temp(hamilton_weather), avg_temp(rotorua_weather), avg_temp(clyde_weather),
                avg_temp(nelson_weather), avg_temp(palmerston_weather), avg_temp(stratford_weather), avg_temp(napier_weather),
                avg_temp(invercargill_weather)))

EV_data = merge(EV_data, HDD_data, by.x = c("year", "month", "weather_region"), by.y = c("Year", "Month", "City"), all.x = T)
EV_data = merge(EV_data, CDD_data, by.x = c("year", "month", "weather_region"), by.y = c("Year", "Month", "City"), all.x = T)
EV_data = merge(EV_data, avg_temp_data, by.x = c("year", "month", "weather_region"), by.y = c("Year", "Month", "City"), all.x = T)

attach(EV_data)



#EV data with columns averaged by month
monthly_EV_data = EV_data[year >= 2017,] %>% 
  group_by(year, month) %>% 
  summarise(mean_kwh = mean(kwh), mean_dist = mean(distance), mean_ef = mean(efficiency))
monthly_EV_data$m = 1:nrow(monthly_EV_data)

#EV data with columns averaged by month and weather region
monthly_reg_EV_data = EV_data[year >= 2017,] %>% 
  na.omit() %>% 
  group_by(year, month, weather_region) %>% 
  summarise(mean_kwh = mean(kwh), mean_dist = mean(distance), mean_ef = mean(efficiency),
            HDD = mean(HDD), CDD = mean(CDD), avg_temp = mean(avg_temp))

```

```{r regions,message = FALSE}
EV_data %>%
  group_by(region, weather_region) %>% 
  summarise(count =  n_distinct(vehicle)) %>% 
  arrange(-count)

```
```{r}
model_pop
```


```{r}
plot(monthly_EV_data$m, monthly_EV_data$mean_ef, type = 'l', xaxt = "n", xlab = "", ylab = "Mean efficiency of EVs (km/kWh)", main = "Time series of EV efficiencies")
axis(1,labels = paste(monthly_EV_data$year, monthly_EV_data$month, sep = "-"), at = monthly_EV_data$m, las = 2,srt = 35)

yearly_line()
```
simple linear model with mean_eff $= t + \ln t + t^2 +$month (as factor). negative squared term means can not use for long term efficiency trend as it will got negative but allows it to better fit the seasonal trend


```{r}
monthly_eff_qm_log = lm(mean_ef ~ m+I(log(m))+I(m^2)+factor(month),data = monthly_EV_data)
summary(monthly_eff_qm_log)
```


```{r}
plot(monthly_EV_data$m, monthly_EV_data$mean_ef, type = 'l', xaxt = "n", xlab = "", ylab = "Mean efficiency of EVs (km/kWh)", main = "Time series of EV efficiencies")
lines(monthly_EV_data$m, predict(monthly_eff_qm_log), col = 'red', lty = 2)

yearly_line()
axis(1,labels = paste(monthly_EV_data$year, monthly_EV_data$month, sep = "-"), at = monthly_EV_data$m, las = 2,srt = 35)
legend("topleft", legend = c("Actual mean EV efficiency", "Modelled prediction EV efficiency"), lty = 1:2, col = 1:2)
```
```{r}
eff_series = ts(monthly_EV_data$mean_ef, frequency = 12)
adf.test(eff_series, alternative = "stationary")
```
we can reject null hypothesis that data is not-stationary. this makes sense as average efficiency should not have significantly changed in a couple of years. use multiplicative instead of additive as preferable to know estimated extra power use? or should i know total extra power used in season?


```{r}
#decomp_eff = decompose(eff_series, "multiplicative")
decomp_eff = decompose(eff_series,"additive")
plot(decomp_eff)
yearly_line(period = 1)
```

```{r}
plot(decomp_eff$figure, type = 'l', main = "Seasonal component of Efficiency of EV", xaxt = "n",
     xlab = "", ylab = "Change in efficiency (km/kWh)")
points(1:12, scale(c(0,monthly_eff_qm_log$coefficients[paste("factor(month)", 2:12, sep = "")]),scale = F),
       type = 'l', col = "red")
axis(1, labels = month.name, at = 1:12, las = 3)
legend("top", legend = c("Seasonal compontent Decompostion", 'Seasonal compontent of linear model'), lty = 1, col = 1:2)
```
will only do for Auckland as too many lines would get crowded

```{r eff_HDD_plot, warning=FALSE}
auck_eff_series = ts(monthly_reg_EV_data$mean_ef[which(monthly_reg_EV_data$weather_region == "Auckland")], frequency = 12)
decomp_auck_eff = decompose(auck_eff_series,"multiplicative")
auck_HDD_series = ts(monthly_reg_EV_data$HDD[monthly_reg_EV_data$weather_region == "Auckland"]/month_length, frequency = 12)
decomp_auck_HDD = decompose(auck_HDD_series,"multiplicative")

par(mar = c(4, 4, 2, 4))
plot(decomp_auck_eff$figure, type = 'b', main = "Auckland Seasonal component Decompostions", xaxt = "n",
     xlab = "", ylab = "Change in efficiency (km/kWh)")
axis(1, labels = month.abb, at = 1:12, las = 3)

par(new=TRUE)
plot(decomp_auck_HDD$figure, type = 'b', col = 2, xlab="", ylab="",axes=FALSE)
mtext("Average Heating Degree (° Days per Day) (Base temp 16°C)",side=4,line=2, cex = 1) 
axis(4)

legend("topright", legend = c("EV efficiency", "HDD"), lty = 1, col = 1:2)
```

intercept base line is Nissan Leaf (24 kWh) 2013-2016
```{r}
eff_lm = lm(efficiency ~ HDD + model, data = EV_data[year >=2017,], na.action=na.omit)
summary(eff_lm)
```
different city weather stations may be measuring colder or warmer regions of the city and therefore may need a slightly different scaling. interesting that Rotorua has lower effect on HDD, could be cause Rotorua is inland Bay of Plenty so its temperature change is more significant that coastal Tauranga which would also be included in bay of plenty.
```{r}
eff_reg_lm = lm(efficiency ~ HDD+ model, data = EV_data[year >=2017,], na.action=na.omit)
summary(eff_reg_lm)
```

https://www.geotab.com/blog/ev-range/

based on this AC should also decrease range. not too obvious in NZ as is kind of cold but in Auckland can see such a trend
what if we include cooling degree days in analysis too? unlike this direct average temp vs efficiency plot this would allow for cooling and heating in the same month that could reduce efficiency. could explain the couple month that have very bad efficiency, possibley have a few cold and warm days but average is nothing unsual 

```{r temp_eff_plot, message=FALSE}
monthly_reg_EV_data[monthly_reg_EV_data$weather_region == "Auckland",] %>% ggplot(aes(avg_temp, mean_ef)) + geom_point() + geom_smooth(method = 'loess')
```

```{r}
eff_h_c_lm = lm(efficiency ~ CDD + HDD +weather_region + model, data = EV_data[year >=2017,], na.action=na.omit)
summary(eff_h_c_lm)
```


```{r lm_plots, fig.show='hide'}
plot(eff_lm)
plot(eff_reg_lm)
plot(eff_h_c_lm)
```
cooling degree days does explain extra variance but not much. likely as not many cooling days above 20 in nz
```{r}
anova(eff_lm, eff_reg_lm, eff_h_c_lm)
```


