---
title: "VKT analysis"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(fig.width=9, fig.height=5) 
library(tidyverse)
library(forecast)
library(tseries)
library(knitr)
rm(list=ls())
```

```{r functions, include = FALSE}
source("functions/yearly_line.R")
source("functions/plot_ts_decomp.R")
```

```{r VKT_data}
vkt = read.csv("downloaded_stats/VKT_main_quart.csv")
vkt_other_yearly = read.csv("downloaded_stats/VKT_other_yearly.csv")
vkt_fuel_yearly = read.csv("downloaded_stats/VKT_fuel_vehicle_type_yearly.csv")
```

```{r VKT_ts}
decomp_vkt = vkt[vkt$year >= 2003,]$Auckland %>% 
  ts(frequency = 4, start = min(2003)) %>% 
  decompose("multiplicative")

plot_decomp(decomp_vkt, ylab = "Million km travelled", title = "Decomposition of Total VKT Time Series")
```


if we make the assumption of the seasonal trend of VKT following a sin wave we can use the equation 
$$y = a \sin{ \left( \frac{2\pi}{12}(x-b)\right)} + c$$
to calculate what such sin wave would have been required to produce the quarterly "Boxes" we can adjust the parameters $a$, $b$, and $c$ to minimize the difference in integral sum. 

$y$ would be a kind of rolling (sum of km)/month

(3 parameters so 3 degrees of freedom and 4 points to base off so will still be slight averaging going on)


```{r find_sin_plot, eval = FALSE}
a_list = seq(0,0.05,0.0001)
b_list = seq(0,12,0.01)
c_list = 1 # it is centered on 1 so does not need to change
#vector order is best a,b,c values and lowest mean squared error
best_comb = c(0,0,0,Inf)
i = 0;

sin_func = function(x, a, b, c) {
  return(a*sin(2*pi*(x-b)/12)+c)
}

for (a in a_list) {
  for (b in b_list) {
    for (c in c_list){
      i = i+1
      quart_sin = c(integrate(sin_func, 0,3, a = a, b = b, c = c)$value, integrate(sin_func, 3,6, a = a, b = b, c = c)$value,
                    integrate(sin_func, 6,9, a = a, b = b, c = c)$value, integrate(sin_func, 9,12, a = a, b = b, c = c)$value)
      error = mean((quart_sin-decomp_vkt_auck$figure*3)^2)
      #should use mean absolute or mean sqaured error
      #error = mean(abs(quart_sin-decomp_vkt_auck$figure*3))
      if (error < best_comb[4]) {
        best_comb = c(a,b,c,error)
      }
    }
  }
}
best_comb
print(i)

```
```{r plot_sin_plot, eval=FALSE}
barplot(decomp_vkt_auck$figure, ylim = c(0.985,1.015), xpd = FALSE, names.arg = c("Q1","Q2","Q3","Q4"), space = 0)
points(x = seq(0,4,0.001), sin_func(seq(0,12,0.003),best_comb[1],best_comb[2],best_comb[3]), type = "l")
```



```{r}

```



